---
title: "R_Assignment2"
author: "Aditya Chandra"
date: "6 December 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading File

The required data is stored in a CSV file named TransportCSOfile.csv and read.csv function can be used to read the data as a dataframe.

```{r transport}
library('stringr')
library('dplyr')
library('ggplot2')
data <- read.csv('data/TransportCSOfile.csv')
str(data)
```

## Renaming Column Headers

Some of the column headers are too long, so we need to rename them.

To rename column Population_Aged_5_Over_By_Means_Of_Travel_To_Work_School_College_On_Foot_2011 (column index 10) to Population_Aged_5_Over_By_Means_Of_Travel_To_Work_School_College_Total_2011 (column index 22), we can split the header into 13 parts and select only the last part as the column's new name.

From column Population_Aged_5_Over_By_Time_Leaving_Home_To_Travel_To_Work_School_College_Before_0630_2011 (column index 23) to Population_Aged_5_Over_By_Time_Leaving_Home_To_Travel_To_Work_School_College_Total_2011 (column index 32), we can rename them by splitting the header name into 15 parts and then concatenating 6th, 7th and 15th part.

From column Population_Aged_5_Over_By_Journey_Time_To_Work_School_College_Under_15_mins_2011 (column index 33) to Population_Aged_5_Over_By_Journey_Time_To_Work_School_College_Total_2011 (column index 40), we can rename them by splitting data into 12 parts and concatenating 6th, 7th and 12th part.

```{r rename}
#To shorten the length of headers from column 10 to 22, split it into 13 parts and take the last part 
for(x in 10:22){
  names(data)[x] <- str_split_fixed(names(data)[x],"_", 13)[,13]
}

#To shorten the length of headers from column 23 to 32, split it into 15 parts and concatenate necessary parts
for(x in 23:32){
  newname <- str_split_fixed(names(data)[x],"_", 15)
  names(data)[x] <- str_c(newname[6],"_",newname[7],"_",newname[15])
}

#To shorten the length of headers from column 33 to 40, split it into 15 parts and concatenate necessary parts
for(x in 33:40){
  newname <- str_split_fixed(names(data)[x],"_", 12)
  names(data)[x] <- str_c(newname[6],"_",newname[7],"_",newname[12])
}

```

## Check and Remove Null Values

First we check the total number of null values in the data.
Since the null values are very less, we can ignore these rows with the help of the function na.omit.

```{r na_removal}
sum(is.na(data))
sapply(data, function(x) sum(is.na(x)))   #To check NA values in each column

#Columns Total_2011 and Time_Leaving_After_0930_2011 have 1 missing value each
#Since there are only two missing values, we can omit them
data <- na.omit(data)

```

## Outlier Treatment

Outlier values can be checked by observing the numerical output of box plot.
Depending on the value of 3rd quartile from summary stats, we can decide upon a upper limit for values in each column and ignore the remaining in order to reduce the number of outliers.

```{r outlier}
#Outliers
summary(data) 
boxplot(data, outline = FALSE)

#Based on the summary stats for each column, we can decide a upper limit for values in each column
#In this way we can reduce the number of outliers
data <-
  subset(data, On_Foot_2011 < 50 &
      Bicycle_2011 < 10 &
      Bus_Minibus_Coach_2011 < 40 &
      Train_Dart_Luas_2011 < 10 & 
      Motorcycle_Scooter_2011 < 5 &
      Car_Driver_2011 < 100 &
      Car_Passenger_2011 < 50 &
      Van_2011 < 20 &
      Other_2011 < 20 &
      Soft_Modes_Comb_2011 < 60 &
      Public_Transport_Comb_2011 < 50 &
      Private_Transport_Comb_2011 < 170 &
      Total_2011 < 300  &
      Time_Leaving_Before_0630_2011 < 20 &
      Time_Leaving_0630_0700_2011 < 30  &
      Time_Leaving_0701_0730_2011 < 30 &
      Time_Leaving_0731_8000_2011 < 50  &
      Time_Leaving_0801_0830_2011 < 50 &
      Time_Leaving_0831_0900_2011 < 60 &
      Time_Leaving_0901_0930_2011 < 30  &
      Time_Leaving_After_0930_2011 < 30 &
      Time_Leaving_Not_Stated_2011 < 15 &
      Time_Leaving_Total_2011 < 250 &
      Journey_Time_Under_15_mins_2011 < 100 &
      Journey_Time_Quarter_To_Under_Half_Hour_2011 < 75 &
      Journey_Time_Half_Hour_To_Under_Three_Quarter_Hours_2011 < 50 &
      Journey_Time_Three_Quarter_Hours_To_Under_One_Hour_2011 < 30  &
      Journey_Time_One_Hour_To_Under_One_Hour_Thirty_Mins_2011 < 20 &
      Journey_Time_One_And_Half_Hours_And_Over_2011 < 10 & 
      Journey_Time_Not_Stated_2011 < 20 & 
      Journey_Time_Total_2011 < 250
  )

```

## Erroneous Value Treatment
An entry in Planning.Region column is named South which should be Southern.

```{r south}
print(unique(data$Planning.Region))
data$Planning.Region[data$Planning.Region == "South"] <- "Southern"
print(unique(data$Planning.Region))

```

## Question 1
Question 1. What is the most popular mode of transport nationally?

Solution - To find out most popular transport mode nationally, we find average number of commuters for each mode and then find the mode that has maximum average.

```{r Question1}
mean_transport_mode <- list()
index <- 10
for(i in 1:10){
  mean_transport_mode[i] <- mean(data[,index])
  index <- index + 1
}
mean_transport_mode

#Most popular transport mode is the one with maximum mean
max_mode <- max(unlist(mean_transport_mode))
print(max_mode)

for(i in 10:20){
  if (mean(data[,i]) == max_mode)
  print(str_c('The most popular mode of transport in Ireland is ', names(data[i])))
}

```

## Question 2
Question 2. How does this (Answer to Q1) compare to the most popular mode of transport in your assigned county?

Solution - To find out most popular transport mode in the assigned county i.e. Donegal, we can make a new dataframe data2 with data related to only the assigned county and repeat the process as previous question.

```{r question2}
data2 <- filter(data, data$County == 'Donegal')    #Filter data related to Donegal only

mean_trans_mode_Donegal <- list()
index <- 10
for(i in 1:10){
  mean_trans_mode_Donegal[i] <- mean(data2[,index])
  index <- index + 1
}
print(mean_trans_mode_Donegal)

max_mode_Donegal <- max(unlist(mean_trans_mode_Donegal))
print(max_mode_Donegal)

#Most popular transport mode is the one with maximum mean
for(i in 10:20){
  if (mean(data2[,i]) == max_mode_Donegal)
    print(str_c('The most popular mode of transport in Donegal is ', names(data2[i])))
}

```

## Question 3
Question 3. What differences are evident between the choice of transportation in the cities compared to the other regions?

Solution - In order to differentiate between choice of transportation between cities and other regions, we divided the data into two dataframes data_cities and data_other, first for cities and second for other regions.Then, we calculate the mean number of commuters for each mode of transport for both the dataframes and generate bar plot for them.

```{r question3}
#For cities
data_cities = data %>% filter(grepl("City",data$County)) %>% summarise_all(funs(mean))
data_cities = t(data_cities[10:21])
data_cities = data.frame(data_cities)
data_cities <- cbind(transportMode = rownames(data_cities), data_cities)

#Plot for transport mode in cities
ggplot(data=data_cities, aes(x=transportMode, y=data_cities)) +
  geom_bar(stat="identity") + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

#For other regions
data_other = data %>% filter(!grepl("City",data$County)) %>% summarise_all(funs(mean))
data_other = t(data_other[10:21])
data_other = data.frame(data_other)
data_other <- cbind(transportMode = rownames(data_other), data_other)

#Plot for transport modes in other regions
ggplot(data=data_other, aes(x=transportMode, y=data_other)) +
  geom_bar(stat="identity") + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

```

## Question 4
Question 4. What proportion of commuters leave home outside of the 8-9am rush hour?

Solution - First calculate the total number of commuters that leave home home for work, school or college. Then calculate the number of commuters that leave outside 8-9 am rush hour by subtracting the the number of commuters between 8-9 am from total number of commuters. Finally, we calculate the proportion of commuters that travel outside the rush hour.

```{r question4}
totalCommuter <- sum(data[,"Time_Leaving_Total_2011"])
print(totalCommuter)

commuter_OutRush <- totalCommuter - sum(sum(data[,"Time_Leaving_0801_0830_2011"]), 
                                        sum(data[,"Time_Leaving_0831_0900_2011"]))

print(commuter_OutRush)
prop <- commuter_OutRush/totalCommuter
print(str_c("Proportion of commuters leaving outside 8-9am is ", round(prop,2)))

```

## Question 5
Question 5. Are commuters in your assigned county likely to travel for longer than 45 minutes each morning?

Solution - First we calculate total number of commuters in Donegal which is the assigned county. Then we calculate the number of commuters who travel more than 45 minutes by adding up number of commuters who travel between three quarters to one hour, one hour to one hour thirty minutes and one hour thirty minutes and over. Then we calculate the ratio of commuters who travel more than 45 minutes to the total number of commuters.

```{r question5}
totalCommuterDonegal <- sum(data2[,"Journey_Time_Total_2011"])
print(str_c("Total number of people who commute in Donegal are ", totalCommuterDonegal))
comm_more_than_45 <- sum(sum(data2[,"Journey_Time_Three_Quarter_Hours_To_Under_One_Hour_2011"]),
                         sum(data2[,"Journey_Time_One_Hour_To_Under_One_Hour_Thirty_Mins_2011"]),
                         sum(data2[,"Journey_Time_One_And_Half_Hours_And_Over_2011"]))
print(str_c("Total number of people in Donegal who travel for more than 45 minutes are ", comm_more_than_45))
prop_exceed_45 <- comm_more_than_45/totalCommuterDonegal
print(str_c("The proportion of people in Donegal who travel longer than 45 minutes is ", 
            round(prop_exceed_45, 3), " which is quite less."))

```

## Question 6
Question 6. How does this (Answer to Q5) compare to other counties in the same NUTS III region?

Solution - First we determine the NUTSIII region for the assigned county (Donegal) which comes out to be Border. Then we create a new dataframe data3, which contains the data only for Border region. Now, we repeat the steps of previous solution for each county in Border region.

```{r question6}
nut <- data2$NUTS_III[1]          #To determine NUTSIII region for assigned county (Donegal)
data3 <- filter(data, data$NUTS_III == nut)   #Filter data for Border NUTSIII region
counties <- data3 %>% distinct(County)

totalCommuterNut <- list()       #List for total number of commuters

for(u in 1:6){
  totalCommuterNut[u] <- sum(filter(data3, data3$County == counties[u,])[,"Journey_Time_Total_2011"])
}
totalCommuterNut

commuterExceed_Nut <- list()       #List for commuters who exceed 45 min
for(county in 1:6){
  commuterExceed_Nut[county] <- sum(sum(filter(data3, data3$County == counties[county,])[,"Journey_Time_Three_Quarter_Hours_To_Under_One_Hour_2011"]),
                       sum(filter(data3, data3$County == counties[county,])[,"Journey_Time_One_Hour_To_Under_One_Hour_Thirty_Mins_2011"]),
                       sum(filter(data3, data3$County == counties[county,])[,"Journey_Time_One_And_Half_Hours_And_Over_2011"]))
}
commuterExceed_Nut

propExceedNut <- list()
for(n in 1:6){
  propExceedNut <- unlist(commuterExceed_Nut[n])/unlist(totalCommuterNut[n])
  print(str_c("Proportion of commuters who travel for more than 45 minutes in ", counties[n,], " is ", 
              round(propExceedNut, 3)))
}

```

## Question 7
Question 7. The residents of which five counties experience the longest commute times?

Solution - First we calculate the total number of commuters who travel for more that one hour thirty minutes in each county and categorise them county-wise. Then we arrange this data in ascending order and select bottom 5 rows which gives us 5 largest values for commuter count. Next, we compare these counts with count for each county and successful matches gives us the names of those 5 counties where commuters experience longest commute time.

```{r}
data_by_county <- data %>% group_by(County) %>% summarise(sum(Journey_Time_One_And_Half_Hours_And_Over_2011))
colnames(data_by_county)[2] <- "Journey_Time_One_And_Half_Hours_And_Over_2011"

top5 <- tail(sort(data_by_county$Journey_Time_One_And_Half_Hours_And_Over_2011),5)

print("The below five Counties experience the longest commute time")

for(i in 1:33){
  for(j in 1:5){
    if(data_by_county$Journey_Time_One_And_Half_Hours_And_Over_2011[i]==top5[j]){
      print(as.character(data_by_county$County[i]))
    }
  }
}

```

## Question 8
Question 8. What proportion of cars used in the morning commute contain only one person?

Solution - First we calculate the total number of car users by combining the columns Car_Driver_2011 and Car_Passenger_2011. Then, we find out the total number of cars where number of passanger is only 1. Finally we calculate the ration of number of cars with one passanger to total number of cars.

```{r question8}
carUser_Total <- data %>% transmute(data$Car_Driver_2011 + data$Car_Passenger_2011)
colnames(carUser_Total)[1] <- "Total Car Users"

one_person_car <- sum(nrow(subset(data,data$Car_Passenger_2011 == 1)),
                    nrow(subset(data,data$Car_Driver_2011 == 1)))

one_person_car_proportion <- one_person_car/sum(carUser_Total)

print(str_c("The proportion of cars used in the morning commute that contains only one person is ", 
      one_person_car_proportion))

```

## Question 10
Question 10. Which Electoral Division within each Planning Region do you propose should be prioritised for investment in public transportation?

Solution - First we determine all the distinct planning regions. Then, within each planning region we calculate the number of public transport users and number of total users for each electoral division. Then we calculate the proportion of number of commuters using public transport to total number of commuters for each electoral division. The electoral division with lowest proportion is the one where we should prioritise investment.

```{r question10}
PlanningRegionList <- data %>% distinct(Planning.Region)
PlanningRegionList <- PlanningRegionList[!apply(PlanningRegionList == "", 1, all),]
PlanningRegionList

for(x in 1:3) {
  sum_public_transport = data %>% 
    filter(Planning.Region == PlanningRegionList[x]) %>%
    group_by(Electoral.Division.Name) %>%
    summarise(sum(Public_Transport_Comb_2011))
  
  sum_total = data %>% filter(Planning.Region == PlanningRegionList[x]) %>%
    group_by(Electoral.Division.Name)  %>%
    summarise(sum(Total_2011))
  public_proportion = sum_public_transport$`sum(Public_Transport_Comb_2011)`/sum_total$`sum(Total_2011)`
  
  print(str_c("For Planning Region ",PlanningRegionList[x], 
              ", it is proposed to invest in public transportation in ", 
              sum_public_transport$Electoral.Division.Name[which.min(public_proportion)],
              " since the proportion of public transport is ", round(min(public_proportion),2), 
              " which is least within this planning region"))
}

```

## Reflection on the Assignment
Overall, the project was an exciting and challenging assignment which demanded good command over string and data manipulation libraries such as stringr and dplyr as well as visualisation library like ggplot2.

One of the many challenges faced while working on this assignment was handling outliers since it was not mentioned how we should deal with them. So, I checked the summary statistics for each column and decided upon a upper cutoff value for data in each column. The upper limit was kept higher than the third quartile for each column in order to retain maximum amount of data without impacting the mean.
Question 9 seemed the most challenging as I was not to come up with a way to tackle it.

It was observed that majority of the people prefer cars and other private modes of transportation for commuting from home to work, school or college compared to other public transportation modes. 

 
